Three Problems for Sophistication

### 2015-01-19 Facticity/Facticity/paper.tex
10 pagina's

\usepackage{amsmath,amsfonts}
\usepackage{color}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\usepackage[mathscr]{eucal}

@@

\begin{abstract}
Kolmogorov complexity is a sound measure of the amount of information in data, but it does not distinguish structural information from random noise. Kolmogorov's definition of the \emph{structure function} was the first attempt to capture this separation. Since then many variations of this idea have been proposed, for which we use \emph{sophistication} as an umbrella term. The multitude of treatments indicates that the idea is considered worth pursuing by multiple distinguished scholars, but at the same time it shows that the issue has never been satisfactorily resolved. We describe three fundamental problems with all existing proposals, showing each of them to be \emph{unsound}. Based on these results we we put forward the view that the problem is fundamental: while many nontrivial and useful structural properties of binary strings can be detected without a priori assumptions, it may be impossible to objectively quantify the sophistication.
\end{abstract}

\section{Introduction}

Kolmogorov complexity breaks away from the tradition in statistics and information theory by allowing the analysis of \emph{individual} binary sequences rather than studying large samples from probability distributions in an assumed class. This is important because for many binary strings, such as a binary encoding of a novel, or a holiday snapshot, or the fossil record, it is impossible to determine what distribution generates such objects. Kolmogorov complexity is often seen as the first step to a statistics without assumptions.

There is, however, a nagging discrepancy between Kolmogorov complexity and what most people would call ``complex''. For example, a sequence of a million coin flips is exceedingly unlikely to have a Kolmogorov complexity of much less than one million bits, even though there is nothing complex about flipping a lot of coins---it's just hard work. For this reason, many scholars have defined alternative complexity measures in the spirit of Kolmogorov complexity, aimed at quantifying not \emph{all} information in a binary string, but only the \emph{meaningful} information in the data. This tradition started with a lecture by Kolmogorov in Talinn in 1973, but many alternative definitions of meaningful information, or \emph{sophistication},\footnotemark were proposed, see the related work section below. In this paper, we investigate three problems with these definitions, and show that they are unsatisfactory. We believe the problem is fundamental, and provide several intuitive arguments why these attempts to define sophistication must fail.

\footnotetext{Most proposals coin their own name along with a specific treatment. For the sake of clarity, we have chosen to use \emph{sophistication} as an umbrella term to refer to all ideas that fit this mold.}

The Kolmogorov complexity of a binary string $x$ is, roughly, the length of the shortest computer program that will output $x$. This length depends on which programming language is used, but only by a constant, independent of $x$. This means that for sufficiently complex objects, the choice of programming language becomes irrelevant and Kolmogorov complexity becomes an \emph{objective} measure of the complexity of binary strings. We can see it as \emph{a property of the data}.

A definition of sophistication $S(x)$ in the spirit of Kolmogorov complexity should have similar robustness guarantees if it is to be interpreted as an objective property of a binary string. We are interested in a measure satisfying the following requirements:

\begin{enumerate}
\item Since it measures information, the sophistication should count the number of bits required for an effective description of the structural properties of a binary string.
\item An analogue of the Invariance Theorem should hold for sophistication: at the very least, there should be strict limits on the degree to which a change in the programming language can affect the sophistication.
\item There should be no constant $c$ such that $S(x)\le c$ for every input $x$. If sophistication is bounded, then knowing its value under one programming language provides no constraints on its value under another language, even for inputs of very high Kolmogorov complexity. Moreover, a bounded sophistication would be at odds with the intuition that there is no limit to the amount of structure that a binary string can exhibit.
\item Similarly, there should be no constant $c$ such that $K(x)-S(x)\le c$ for all $x$, because then sophistication would be equivalent to Kolmogorov complexity. 
\end{enumerate}

For most proposed definitions of sophistication, we can prove that they fail one or more of the conditions above. Vit\'anyi's definition \cite{vitanyi2004meaningful} is the one exception that we cannot \emph{prove} to conflict with our requirements, but we can show that only extremely deep strings (i.e., strings that require an enormous amount of processing to construct) yield a large sophistication, while intuitively it seems there must be objects that contain a lot of structure all of which can be described without much calculation. Moreover in our discussion section we provide arguments that suggest that it should be possible to generalise our negative results to show that Vit\'anyi's sophistication is not invariant.

A valid definition of sophistication must contend with the following issues. First, while all definitions are based on a two-part encoding of the data sequence, some researchers use Kolmogorov complexity as a \emph{building block} of their proposed code, while others look for the two parts of the code \emph{inside} the Kolmogorov complexity: when the definition of Kolmogorov complexity is opened up, it turns out to consist of two parts itself. In Section~\ref{section:indices}, we show how this second approach sets us up to suffer from to the so-called \emph{nickname problem}: the sophistication becomes highly dependent on the chosen programming language. This immediately breaks invariance, unless additional requirements are made on the language. We also show that there do exist languages (efficient enumerations of the partial recursive functions) that avoid this issue.

The second issue is that of \emph{overfitting}, which we consider in Section~\ref{section:overfitting}. Overfitting is a common problem in the statistical literature, that refers to the tendency of some model selection procedures to select a very complex model that provides a very good fit to the observed data, but does not generalise well to unseen data. For example, in a polynomial regression setting, a higher order polynomial can always provide better fit to a data set. Therefore, even if the data are in fact noisy samples from a parabola, naive model selection methods may select a much higher order polynomial that provides a worse description of the trend in the data, encoding all noise in the model parameters. The same can occur for the sophistication. We somehow need to make sure that the two-part representation that determines the sophistication does not store noise in the ``structure'' part of the code. In statistics, this is typically addressed by penalising complex models. However, in the sophistication setting, such penalties tend to break the delicate balance between structural information and noise, and lead to the opposite problem: underfitting.

Underfitting (Section~\ref{section:underfitting}) is the opposite problem, where the selected model is simple, but fails to represent part of the structure in the data. It is an especially serious problem for sophistication because the models for used in sophistication are so powerful. In particular, in any programming language, there are programs that implement an interpreter for \emph{another} language. Such interpreters, or \emph{universal models}, can be described in a constant amount of bits, and can generate the data with from a program the length of the Kolmogorov complexity, essentially encoding all information as noise. If complex models are penalized to avoid overfitting, then the problem becomes to make sure that universal models are not \emph{always} preferred for complex data. The usual workaround is to restrict the set of allowed models to exclude the universal ones. For example, the model class may be restricted to the set of all total recursive functions (only programs that halt for all inputs) rather than all partial recursive functions (all programs). While this patches the problem of universal models to an extent, it is questionable whether it adequately solves the problem of underfitting in general.

Finally, in the discussion in Section~\ref{section:conclusion} we argue that while two-part coding can yield useful insights into the structure of the data and allows us to identify some models as unlikely to capture all structure in the data, it is probably not possible to uniquely separate structure from noise and identify a \emph{single} model as ``best'': in general many models, of vastly different complexities, may be reasonable models of the data generating mechanism. Rather than doggedly trying to ``fix'' this property of algorithmic statistics, we propose embracing the idea that sometimes the data allows for multiple equivalent interpretations, and that there is no such thing as sophistication.

\subsection{Related work}
The idea of separating structure from noise originated with Kolmogorov's lecture in Tallinn, Estonia in 1973 introducing the \emph{structure function} \cite{cover1985kolmogorov}. In this theory, the two-part description of the data consists of a program that generates a finite set, and an index of the original object in the set. The length of the program then measures the structure in the data. This approach was discussed more extensively, and generalised, by Vereshchagin, G\'acs and Vit\'anyi \cite{vereshchagin2004kolmogorov,gacs2001algorithmic}.

Koppel \cite{koppelSoph1988,koppel1991almost} also built on Kolmogorov and Cover's work to define a measure of the meaningful information in a binary string, called \emph{sophistication}, the term we adopt in this paper. He changed the class of models from finite sets to total functions and used monotone functions, instead of prefix-free Kolmogorov complexity, to be able to extend the theory to infinite strings, an approach which has been abandoned by subsequent authors. We will only consider the finite version (based on regular functions) in this paper.

In 1996 Gell-Mann and Lloyd introduced \emph{effective complexity} \cite{gellmann1996information}. Working from the perspective of statistical mechanics, they arrived (largely independently, it seems) at a notion that fits the mold of sophistication.

Antunes et al. \cite{antunes2009sophistication} note that the constant cutoff used by Koppel to separate candidate two-part descriptions from those deemed to be too inefficient, is a source of instability, a result in the spirit of this paper. They propose a different definition called \emph{coarse sophistication}. As we shall see, this version behaves very differently from regular sophistication, and is closer to computational depth. Vereshchagin \cite{vereshchagin2013algorithmic} proposes an alternative solution to the instability issue.

Adriaans' \emph{Facticity}, introduced in 2012\cite{adriaans2012facticity}, relaxes the model class to include all partial recursive functions. Adriaans also identifies the nickname problem discussed in Section~\ref{section:indices}.

\subsection{Preliminaries, definitions and desiderata}

Let $\B = \{0,1\}^*$. We deal with partial functions $\phi: \B \to \B$, which we will also call \emph{models}. Such functions are \emph{computable} if they can be computed by a Turing machine as described in \cite[Definition~1.7.1]{li1993introduction}. A function is prefix-free if its domain, $D_f = \{y : f(y) \neq \infty\}$ is a prefix set. A function $f$ is \emph{total} if $D_f = \B$. A prefix-free function is \emph{complete} if every infinite binary string has a member of $D_f$ as a prefix. \footnotemark

\footnotetext{Completeness is analogous to totality: a total function corresponds to a Turing machine that halts for all its inputs, while a complete function corresponds to a Turing machine that always halts, so long as there are enough bits on the input tape.}

\begin{definition}[Numberings]
A numbering is an enumeration of of the partial computable functions, denoted as $\phi_1, \phi_2, \ldots$. We fix one canonical, which we know to be an effective description, that is for any $i \in \N, y\in \B$, we can effectively compute $\phi_i(y)$, where the computation diverges if $\phi_i(y) = \infty$. We call a numbering $\num{\psi}$ \emph{acceptable} if there exist total, computable functions $a, b: \N \to \N$ such that for all $i$, $\phi_i = \psi_{b(i)}$ and  $\psi_i = \phi_{b(i)}$. We use the greek letters without a subscript (eg. $\phi$ and $\psi$) to refer to the numbering as a whole.
\end{definition}

\begin{definition}[Prefix encoding function]
  A prefix encoding function $f:\B\to\B$ is an injective
  function that maps the binary strings to a prefix-free set. There
  are prefix encoding functions $g$ with the property that
  $|g(x)|\le|x|+2\log(|x|+2)|$ \cite{li1993introduction}. We fix such a function and
  denote it by $\br{x}=g(x)$.
\end{definition}

\begin{definition}[Complexity]\belowdisplayskip=-12pt
Let $\phi_1,\phi_2,\ldots$ be the canonical acceptable enumeration of $\C$. Let $\Pi$ be a computable subset of the natural numbers such that  $\{\phi_i:i\in\Pi\}$ corresponds to the set of all computable prefix functions. We define the plain and prefix Kolmogorov complexity as 
\begin{align*}
C(x)&=\min\{|\bar\imath y|:\phi_i(y)=x\},\\
K(x)&=\min\{|\bar\imath y|:\phi_i(y)=x, i\in\Pi\}.\\
\end{align*}\label{definition:complexity}\vspace{-18pt}
\end{definition}
This definition of $K$ is uncommon. It is equivalent to the standard definition as used in \cite{li1993introduction}, but this form is more convenient for defining the $K$-complexity of non-prefix functions (see Definition~\ref{definition:model-complexity} below).

We will generalize the main proposals for sophistication into a single notation. We start with a class of models, we consider all two-part descriptions of the data using these models, we select the set of representations with the shortest code-lengths and we select the smallest model from this set. The size of the smallest model represents the sophistication of the data.

\begin{definition}[Models and model classes] A \emph{model class} is a set of models. We distinguish four important model classes:
  \begin{itemize}
  \item $\C$ is the set of all partial computable functions,
  \item $\K\subset\C$ is the set of all prefix-free
    functions,
  \item $\T\subset\C$ is the set of all total functions,
  \item $\P\subset\K$ is the set of all complete, prefix-free functions,
  \item $\F\subset\P$ contains, for every finite set $S$, a computable
    surjective function mapping the binary sequences of length
    $\lceil\log S\rceil$ onto $S$. In other words, this is the model class of finite sets.
  \end{itemize}
\end{definition}
We note that classes $\T$ and $\P$ are unusual in that they are not enumerable.
\begin{definition}[Representations and code-lengths]
A pair $(m,y)$ is a \emph{representation} for $x$ if $m$ is a model and $m(y)=x$.
Define the following code-length functions for the representations:
\begin{itemize}
\item $L(m)$ is the length of a self-delimiting description of
  the model $m$, i.e. it is the length function of a prefix code for models. 
\item $L^m(x)=L(m)+\min\{|y|:m(y)=x\}$ measures the number of bits required to describe both the model $m$ and the data $x$ using the model.
\item $L^\M(x)=\min\{L^m(x):m\in\M\}$ measures the number of bits
  required to describe $x$ using the best model in the model class.
\end{itemize}
\end{definition}
We can now define which of these representations express the data most efficiently:
\begin{definition}[Candidate set]
The \emph{candidate models} for $x$ from model class $\M$ and threshold function $\tau$ are
\[
  C^\M_\tau(x):=\{m\in\M:L^m(x)\le \tau(x) + c\}
\]
with $c$ a constant independent of $x$ which may be chosen arbitrarily. The default threshold function is $\tau^\circ(x)=L^\M(x)$. We will often omit the subscript if the default threshold function is used. 
\end{definition}

\begin{definition}[Sophistication]\belowdisplayskip=-12pt
  The model complexity, meaningful information, effective complexity, facticity or \emph{sophistication},  of a string $x$ given a model class $\M$ and threshold function $\tau$ is
  \[
  \s_\tau^{\M}(x):=\min\{L(m):m\in C^\M_\tau(x)\}\p
  \]\label{definition:sophistication}
\end{definition}
Finally, we must formalize what we mean when we say that we expect sophistication to be invariant. It may be that a less strict invariance than we have for the Kolmogorov complexity, still yields a useful sophistication.
\begin{definition}
Let $\gamma:\N\to \N$ be some nondecreasing function. Two functions $f$ and $g$ are $\gamma$-equivalent if $|f(x) - g(x)| \in O(\gamma(\max(f(x), g(x)))$.
\end{definition}
We say that a formalization of sophistication $\s_\phi$ built on acceptable numbering $\phi$ is $\gamma$-invariant to the choice of numbering is $s_\phi$ and $s_\psi$ are $\gamma$-equal for any two acceptable numberings $\phi$ and $\psi$. 
\begin{lemma}
If a function $\S_\phi(x)$ is bounded for a given numbering, it is either not $\gamma$-invariant or it is  $\gamma$-equal to a bounded function.
\end{lemma}
The proof follows directly from the definitions. Most lemmas in the rest of the paper will simply prove that a given variant is bounded for a specific numbering, implicitly using this lemma to show that the variant is unsound.

\section{Inefficient indices}
\label{section:indices}
We start from a simple intuition: the definition of Kolmogorov complexity already uses a description of the data in two parts; a computable function and an input. Can we assume that, when such a description is as short as the Kolmogorov complexity, it automatically creates a good separation into the structural part of the data and the noise? As we will show, the representation of function plays a fundamental part in answering this question.

The issues can be very subtle and require a careful attention that is rarely given in the literature. We will first illustrate what happens if any acceptable numbering is allowed:

\begin{lemma}
Let $\phi$ be some acceptable numbering, and $\s_\phi(x)$ be any function which gives the length of some index $i$ such that $\phi_i(y) = x$ for some $y$.

No such function can be both $\gamma$-invariant and unbounded for any $\gamma(n) \in o(n)$.
\end{lemma}
\begin{proof}
Choose some $\gamma(n) \in o(n)$. Assume towards contradiction that the sophistication is $\gamma$-invariant and unbounded.

>>>>

\usepackage{amsmath,amsfonts}
\usepackage{color}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\usepackage[mathscr]{eucal}
\usepackage{thmtools}

@@

\begin{abstract}
Kolmogorov complexity measures the amount of information in data, but does not distinguish structure from noise. Kolmogorov's definition of the \emph{structure function} was the first attempt to measure only the structural information in data, by measuring the complexity of an objectively best model under two-part coding. Since then, many variations of this idea have been proposed, for which we use \emph{sophistication} as an umbrella term. We describe three fundamental problems with all existing proposals, showing each  to be \emph{unsound}. Consequently, we put forward the view that the problem is fundamental, it may be impossible to objectively quantify the sophistication.
\end{abstract}
\section{Introduction}

Komogorov complexity gives us a sound definition of the amount of information contained in a bitstring. There is, however, a nagging discrepancy between Kolmogorov complexity and what most people would consider complexity. For example, a sequence of a million coin flips will almost certainly have maximal Kolmogoroc complexity, even though there is nothing complex about flipping a lot of coins---it's just hard work. For this reason, many scholars have defined alternative complexity measures in the spirit of Kolmogorov complexity, aimed at quantifying not \emph{all} information in a binary string, but only the \emph{meaningful} information in the data. This tradition started with a lecture by Kolmogorov in Talinn in 1973, but many alternative definitions of meaningful information, or \emph{sophistication},\footnotemark were proposed. In this paper, we investigate three serious problems with these definitions. We believe that ultimately, some of these issues are insurmountable, and provide several intuitive arguments why any attempt to define sophistication must fail.

The Kolmogorov complexity of a binary string $x$ is, roughly, the length of the shortest computer program that will output $x$. This length depends on which programming language is used, but only by a constant independent of $x$. This property is known as \emph{invariance}. This means that for sufficiently complex objects, the choice of programming language becomes irrelevant and we can see the Kolmogorov complexity  as \emph{a property of the data}.

A definition of sophistication $S(x)$ in the spirit of Kolmogorov complexity should have similar robustness guarantees. We are interested in a measure satisfying the following requirements:

\begin{enumerate}
\item Since it measures information, $S(x)$ should count the bits required for an effective description of the structural properties of a binary string.
\item An analogue of invariance should hold for sophistication: at the very least, there should be strict limits on the degree to which a change in the programming language can affect the sophistication.
\item There should be no constant $c$ such that $S(x)\le c$ for every input $x$. If sophistication is bounded, then knowing its value under one programming language provides no constraints on its value under another language. Moreover, a bounded sophistication would be at odds with the intuition that there is no limit to the amount of structure that a binary string can exhibit.
\item Similarly, there should be no constant $c$ such that $K(x)-S(x)\le c$ for all $x$, because then sophistication would be equivalent to Kolmogorov complexity. 
\end{enumerate}
For most proposed definitions of sophistication, we can prove that they fail one or more of the conditions above. Vit\'anyi's definition \cite{vitanyi2004meaningful} is the one exception that we cannot \emph{prove} to conflict with our requirements, but we can show that only strings that require an enormous amount of processing to construct, yield a large sophistication.

A valid definition of $S$ must contend with the following issues. First, the representation of the model is crucial. This is particularly important to those treatments that `open up' the definition of $K$, to consider the representation that witnesses the Kolmogorov complexity. In Section~\ref{section:indices}, we discuss the \emph{nickname problem}: the sophistication becomes highly dependent on the chosen programming language, unless it can represent functions with maximum efficiency.

The second issue is that of \emph{overfitting}, which we consider in Section~\ref{section:overfitting}. Overfitting is a common problem in the statistical literature, that refers to the tendency of model selection to choose a very complex model that provides a very good fit to the observed data, but does not generalise well to unseen data. For example, in a polynomial regression setting, a polynomial of sufficiently high order can always cover all data points. However, if the data was sampled from a straight line, with a small amount of noise, such a polynomial would be far too complex: it would encode all noise inside the model. The same can occur for the sophistication. We somehow need to make sure that the two-part representation that determines the sophistication does not store noise in the ``structure'' part of the code. In statistics, this is typically addressed by penalising complex models. In sophistication, however, such penalties tend to break the balance between structural information and noise, and lead to the opposite problem: underfitting.

Underfitting (Section~\ref{section:underfitting}) occurs when the selected model is simple, but fails to represent some of the structure present in the data. In particular, in any programming language, there are programs that implement an interpreter for \emph{another} language. Such \emph{universal models} can be described in a constant amount of bits, and  represent the data with a program the length of the Kolmogorov complexity, essentially encoding all information as noise. If complex models are penalized, then the problem becomes to make sure that universal models are not \emph{always} preferred for complex data. The usual workaround is to restrict the set of allowed models to exclude the universal ones. For example, the model class may be restricted to the set of all total recursive functions. While excludes universal models, it is questionable whether it adequately solves the problem of underfitting in general.

Finally, in the discussion in Section~\ref{section:conclusion} we argue that while two-part coding can yield useful insights into the structure of the data and allows us to identify some models as poor representations, it is probably not possible to uniquely separate structure from noise and identify a \emph{single} model as ``best'': in general many models, of vastly different complexities, may be reasonable representations. Rather than doggedly trying to ``fix'' this property of algorithmic statistics, we propose embracing the idea that the data allows for multiple equivalent interpretations, and that there is no such thing as sophistication.

\subsection{Related work}
The idea of separating structure from noise originated with Kolmogorov's \emph{structure function} \cite{cover1985kolmogorov}. In this theory, the data is described as finite set together with an index in the set. The complexity of the set determines the sophistication. This approach was discussed more extensively, and generalised in \cite{vereshchagin2004kolmogorov,gacs2001algorithmic}.

Koppel \cite{koppelSoph1988,koppel1991almost} also built on the structure function work to define a measure of the meaningful information in a binary string, coining \emph{sophistication}, the name we adopt in this paper.

In 1996 Gell-Mann and Lloyd introduced \emph{effective complexity} \cite{gellmann1996information}. Working from the perspective of statistical mechanics, they arrived (independently, it seems) at a notion that fits the mold of sophistication.

Antunes et al. \cite{antunes2009sophistication} note that the constant cutoff used by Koppel is a source of instability, a result in the spirit of this paper. They propose a different definition called \emph{coarse sophistication}. As we shall see, this version behaves very differently from regular sophistication, and is closer to computational depth. Vereshchagin \cite{vereshchagin2013algorithmic} proposes an alternative solution to the instability issue. Adriaans' \emph{Facticity}, introduced in 2012\cite{adriaans2012facticity}, relaxes the model class to include all partial recursive functions. Adriaans also identifies the nickname problem discussed in Section~\ref{section:indices}.

\subsection{Notation}

Let $\B = \{0,1\}^*$. We deal with computable partial functions $\phi: \B \to \B$, which we will call \emph{models}. A function is prefix-free if its domain, $D_f = \{y : f(y) \neq \infty\}$ is a prefix free set. A function $f$ is \emph{total} if $D_f = \B$. A prefix-free function is \emph{complete} if every infinite binary string has a member of $D_f$ as a prefix. We fix a prefix code with $|\bar{x}| \le|x|+2\log(|x|+2)|$ (eg. \cite[Section~1.4]{li1993introduction}).

\begin{definition}[Numberings]
A numbering is an enumeration of of the partial computable functions, denoted as $\phi_1, \phi_2, \ldots$ or simply $\phi$. We fix one effective numbering $\phi$. We call a numbering $\num{\psi}$ \emph{acceptable} if there exist total, computable functions $a, b: \N \to \N$ with $\forall: i$, $\phi_i = \psi_{b(i)}$ and  $\psi_i = \phi_{b(i)}$.
\end{definition}

\begin{definition}[Complexity]
Let $\Pi$ be a computable subset of the natural numbers such that  $\{\phi_i:i\in\Pi\}$ corresponds to the set of all computable prefix functions. We define the plain and prefix Kolmogorov complexity as $C(x)=\min\{|\bar\imath y|:\phi_i(y)=x\}$ and $K(x)=\min\{|\bar\imath y|:\phi_i(y)=x, i\in\Pi\}$ respectively.
\label{definition:complexity}
\end{definition}

\begin{definition}[Model classes] A \emph{model class} is a set of models. We distinguish four important model classes: the partial (computable) functions ($\C$), the prefix functions ($\K \subset \C$), the total functions ($\T \subset \C$), the complete functions ($\P \subset \K$) and the finite sets ($\F \subset \P$, encoded as a prefix-free function mapping all strings of length $\lceil\log S\rceil$ to members of $S$).
\end{definition}

\begin{definition}[Representations]
A pair $(i,y)$ is a \emph{representation} for $x$ if $\phi_i(y)=x$.
Define the following code-length functions for the representations:\\
\-\hspace{1cm}$L(m)$ is the length of a self-delimiting description of the model $m$.\\ 
\-\hspace{1cm}$L^m(x)=L(m)+\min\{|y|:m(y)=x\}$\\
\-\hspace{1cm}$L^\M(x)=\min\{L^m(x):m\in\M\}$
\end{definition}
We can now define sophistication generically:
\begin{definition}[Sophistication]
The \emph{candidate models} for $x$ from model class $\M$ and threshold function $\tau$ are
$C^\M_\tau(x):=\{m\in\M:L^m(x)\le \tau(x) + c\}$
with $c$ a constant independent of $x$ which may be chosen arbitrarily. The default threshold function is $L^\M(x)$. The \emph{sophistication} is \\ 
\-\hspace{1cm}$\s_\tau^{\M}(x):=\min\{L(m):m\in C^\M_\tau(x)\}$\label{definition:sophistication}
\end{definition}
Finally, we generalize the notion of invariance to allow for less strict versions. 
\begin{definition}[$\gamma-invariance$]
Let $\gamma:\N\to \N$ be some nondecreasing function. Two functions $f$ and $g$ are $\gamma$-equivalent if $|f(x) - g(x)| \in O(\gamma(\max(f(x), g(x)))$. We say that a function $\s_\phi(x)$ is \emph{$\gamma$-invariant} to the choice of numbering if $s_\phi$ and $s_\psi$ are $\gamma$-equal for any two acceptable numberings $\phi$ and $\psi$. 
\end{definition}

\begin{lemma}
If a function $\s_\psi(x)$ is bounded for some $\psi$, it is either not $\gamma$-invariant or it is  $\gamma$-equivalent to a bounded function.
\end{lemma}

\section{Inefficient indices}
\label{section:indices}
We start from a simple intuition: the definition of $K$ already uses a description of the data in two parts; a computable function and an input. Can we assume that, when such a description is optimal, it automatically separates the structural and random information? As we will show, the representation of functions plays crucial fundamental.

We will first illustrate what happens if any acceptable numbering is allowed:
\begin{restatable}{lemma}{ineffindices}
Let $\phi$ be some acceptable numbering, and $\s_\phi(x)$ be any function which gives the length of some index $i$ such that $\phi_i(y) = x$ for some $y$.

No such function can be both $\gamma$-invariant and unbounded for any $\gamma(n) \in o(n)$.
\end{restatable}
\noindent In the proof of this lemma (Section~\ref{section:appendix-inefficient-indices}), we deliberately used ridiculously inefficient index functions. We could use a more efficient index, but most definitions of sophistication do not mention any such restrictions. As we shall see, any inefficiency in the index will lead to a lack of invariance in sophistication.

The treatments that fall prey to this problem are Koppel and Atlan's \cite{koppelSoph1988,koppel1991almost},  and several of its variants \cite{antunes2009sophistication,antunes2013sophistication}.

Interestingly, inefficient indices are no problem in the definition of Kolmogorov complexity. However, this robustness does \emph{not} carry over to definitions of sophistication. Simply put, the length of a model in an ordinary description does not represent its complexity, much like the length of a compressible string does not reflect its information content. In order to obtain a robust model complexity, we use an extension of Kolmogorov complexity:

\begin{definition}[\cite{grunwald2004shannon,vitanyi2004meaningful}]
The complexity of a partial recursive function is defined by $C_\phi(f) = \min\{C(i):\phi_i=f\}$ and $K_\phi(f) = \min\{K(i):\phi_i=f\}$.\label{definition:model-complexity}
\end{definition}
We show in the appendix (Lemma~\ref{lemma:invariance}) that these are invariant, ie. complexity is a property of a function. There are two ways to use this notion of model complexity for more robust attempts to define sophistication. Confusingly, both approaches are used in the literature.
\begin{enumerate}
  \item We maintain that the representation witnessing the Kolmogorov complexity determines the sophistication, but impose the additional constraint that an efficient, acceptable numbering is used. Specifically, that the length of the indices satisfies $|i| = C(\phi_i)$. This approach is taken by Adriaans \cite{adriaans2012facticity}.
  \item On the other hand, rather than distinguishing the two components of the code \emph{within} the definition of Kolmogorov complexity, we can also use $K$ to measure the complexity of the \emph{model} (as in Definition~\ref{definition:model-complexity}), which is then the first part of a two-part code describing the data. With this approach, the total code length is $K(\phi_i)+|y|$, with $\phi_i(y)=x$. In this approach, we can construct everything on the basis of a single acceptable numbering, without additional constraints. This approach is used in  \cite{cover1985kolmogorov,gacs2001algorithmic,vitanyi2004meaningful,gellmann1996information}.
\end{enumerate}
To show that the two perspectives are equivalent, we must show that there exists an acceptable numbering which is efficient enough:
\begin{definition}[Faithful Numberings]\label{def:faithful}
  A numbering $\num{\psi}$ is \emph{faithful} if
  for all $i$ there is a $j$ such that $\psi_i=\psi_j$ and
  $|j|\le C(\psi_j)+c$, for some constant $c$.
\end{definition}
This definition is a refinement of \cite[Definition~10]{adriaans2012facticity}. We show in the appendix (Lemma~\ref{lemma:faithful-numberings}) that such numberings exist. Unfortunately, even with a faithful numbering, we can fail:
\begin{restatable}{lemma}{ineffprefix}
Let $\phi$ be faithful and let $L(\phi_i) = |\bar\imath|$, for some prefix encoding function. If $\min\{i\ge i_0:|\bar\imath|-K(i)\}$ is an unbounded function of $i_0$, then $\s^\C$ is bounded for all numberings.\label{lemma:prefix-inefficiency}
\end{restatable}
\noindent This result applies to \cite{adriaans2012facticity}. To avoid the problem, the model $\phi_i$ should be encoded in $K(\phi_i)$ bits, relying on $K$'s own prefix encoding to delimit the model.

Note that the condition for this lemma likely applies to all possible computable  prefix encoding functions (see Section~\ref{section:prefix-inefficiency}). Thus, even if we have a faithful numbering, we can only use it if we don't count the cost of delimiting the model from the input (as in \cite{koppelSoph1988}). In the following, we will take the second perspective: the representations are two-part descriptions $(i, y)$ such that $\phi_i(y)=x$, with codelength $K(\phi_i) + |y|$. 
\section{Overfitting: the singletons}
\label{section:overfitting}
We now come to the main issue plaguing sophistication. If we use model class $\K$, the candidate set always contains two representations that do not provide a good separation. The first is a \emph{universal model}: a basic universal Turing machine. The second is a \emph{singleton model} a model which returns $x$ for any input. As we will show, in many settings, we can ensure that one of them is the \emph{only} model in the candidate set. The following lemma generalizes the principle:

\begin{restatable}{lemma}{coolone}
\label{lemma:thecoolone}
  Let $\psi_1,\psi_2,\ldots$ be any acceptable enumeration of the partial recursive functions.
  Let $\M$ be any model class, let $\X$ be any set of binary sequences and let $D:\{0,1\}^*\to\N$ be a computable decoding function with a prefix-free domain that maps function descriptions to their indices in $\psi$, i.e. if $f=\psi_{D(p)}$ then $p$ is a $D$-description of $f$. Let $\M'=\{\psi_i:i\in\tn{range}(D)\}$. Further assume there is a constant $c$ such that:\\
\-\hspace{1cm}(1)$\forall_{f\in\M'}:\min\{|p|:\psi_{D(p)}=f\}\le K^\psi(f)+c$\\
\-\hspace{1cm}(2)$\forall_{x\in\X}:L^{\M',\psi}(x)-L^{\M,\psi}(x)\le c$.\\
Then there is an enumeration $\phi_1,\phi_2,\ldots$ of the partial recursive functions such that $S^\phi_{\M}(x) = |\bar 0|+S^\psi_{\M'}(x)$ for all $x\in\X$.
\end{restatable}
\noindent We can now show numberings for which the singletons are always selected:

\begin{restatable}[Overfitting]{lemma}{overfitting}
Let $\M \subseteq \K$ be a model class where for every $x\in\X$ there is a singleton model $f\in\M$ with $f(\epsilon)=x$. Then there is an enumeration $\phi_1,\phi_2,\ldots$ of the prefix partial recursive functions, and a constant $c$, such that
\[
K(x)-S^{\M,\phi}(x)\le c
\]
for all $x\in\X$.
\end{restatable}
\noindent In short, this lemma tells us that there are numberings for which the singleton models always determine the sophistication. This means that for any dataset, the sophistication is equal to the Kolmogorov complexity.

Many variants in the literature (including the structure function) are susceptible to this problem \cite{cover1985kolmogorov,gacs2001algorithmic,vereshchagin2013algorithmic}. \cite{vitanyi2004meaningful,adriaans2012facticity} avoid it by not using self-delimiting representations. Since a singleton representation places all information in the (self-delimiting) model part, singletons are automatically not optimally efficient. While this dodges the issue of the singletons, it does not show that $S$ is invariant.

Another approach may be to make the constant threshold dependent on the choice of enumeration, although we are not aware of any published efforts.

\section{Underfitting: the universal model}
\label{section:underfitting}
On the other end of the spectrum, there exist representations with all information in the noise, using a universal model. We can easily show that if such models exist, we can choose our numbering so that they are always chosen.
\begin{restatable}[underfitting]{lemma}{underfitting}
Let $\M$ be a model class containing model $u$, called a \emph{universal model} with $\exists c \forall m \in \M : L^u(x) \leq L^m(x) + c \p$. Then, there exist numberings such that $\s^\M$ is bounded.
\end{restatable}
\noindent Most sophistication measures avoid the underfitting problem by limiting their model class, usually to either finite sets ($\F$) or total functions ($\T$ or $\P$). This eliminates the universal model, but as we shall see, it does not solve the problem for most data. The variants that are susceptible to the this type of underfitting are facticity \cite{adriaans2012facticity} and some versions of effective complexity \cite{gellmann1996information}. The latter aims to bypass the problem by selecting carefully from the candidate set, but as we have shown, for some numberings the candidate set contains only the universal model.

We have seen that some variants of sophistication sidestep the simplest examples of underfitting by using total models. Does this ensure that these variants are not bounded? A well-known result in algorithmic statistics shows that there exist strings for which no finite set is as good a model as the singleton set. In this section we will focus on the variants $S^\T$ and $S^T_K$ \cite{vitanyi2004meaningful}, as those are the variants most resistant to proofs of non-invariance.

\begin{restatable}{lemma}{nonstochastic}
There are infinitely  $x$ such that $S^\F_K(x) \geq K(x)/5$. \label{lemma:structure-function-is-not-bounded}
\end{restatable}

\begin{restatable}{lemma}{vitanyi}
Let $f$ be a total, computable function, such that for some $d$, $f(d) = x$ and let $k = K(f) + |d|$. Then there exists a finite set $S$ containing $x$ and a constant $c$, such that $K(S) \leq K(f) + K(|d|) + c$ and $\log |S| \leq |d|$.\label{lemma:total-to-sets}
\end{restatable}
\noindent This lemma is a variation on \cite[Lemma~7.2]{vitanyi2004meaningful}. We use this lemma to transport the result of Theorem~\ref{lemma:structure-function-is-not-bounded} to the model class of total functions:

\begin{restatable}{theorem}{dogfood}
There are infinitely many $x$ with $S^\T_K(x) \geq K(x)/6$ and $S^\T(x) \geq K(x)/6$. 
\end{restatable} 
\noindent This tells us that $S^\T_K(x)$ and $S^\T(x)$ can truly grow to a significant part of the total description length. While this shows that sophistication, if defined with particular care, is not entirely vacuous, we can shed some light on exactly which strings end up with high sophistication, using the principle of \emph{depth}. We will use the following definition:
